Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2021-01-27 SIR_manuscript.tex
comments addressed in Method

\usepackage{xspace}
%\newcommand{\Rlogo}{\protect\includegraphics[height=2ex,keepaspectratio]{pix/Rlogo.pdf}\xspace}
% BMB: we don't need the weird graphical logo ...
\newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}

@@

We developed a deterministic model, \crefrange{eq1}{eq12}, which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered (thus this is an SIR-type model), and testing status categorizes people as \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (Figure~\ref{fig:flowchart}) \bmb{make sure we're using consistent terminology for the testing compartments}. Two `accumulator' compartments, $N$ and $P$, were also incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of $\Rnum$ are presented in the Supplementary Materials.

\bmb{need a little bit more detail moved from the Supp. (1) If equations are in the supp we need to tag the equation numbers (e.g. eq S1-S14); (2) we might want the whole set of equations to have a single number (we could also sublabel S1(a), S1(b) etc., but as we never refer to the individual equations we probably don't need to bother); (3) at the very least, $\Lambda$ should be defined in the body of the text (we need it to understand the figure), and $\beta$ should be included in the parameter table}

\begin{figure}[!h] 
\begin{center} 
\includegraphics[scale=0.3]{./pix/sir_comp.png}
\caption{\small Flowchart of the SIR (Susceptible-Infectious-Recovered) model, \crefrange{eq1}{eq12}. For the description of the parameters see \cref{tab:params}.
\label{fig:flowchart}}
\end{center} 
\end{figure}

\jd{It would be nice to modularize the model somehow, but probably not worth anyone's time.}
\bmb{Can you say what you mean by this? Just curious.}
\jd{I wrote this when the equations were here; I meant it might be good to use less equations, by writing summaries of the SIR process and the within-each-box testing process.}

Table~\ref{tab:params} defines the model parameters, which are generally straightforward
\emph{per capita} flows between compartments, or modifiers to these flow rates.
The novel component of the model comes in through the compartment-specific relative testing weights $W_S$, $W_I$ and $W_R$; these give the relative rates at which people in the $S$, $I$, and $R$ compartments are tested, respectively. For example, $W_I/W_S=2$ means that infected individuals are tested at twice the \emph{per capita} rate of susceptible individuals.

In order to link to more applied models, we constructed this model so we could specify  to be able to specify the total \percap testing rate. We do this by defining the weighted size of the testing pool $W = W_S S_u + W_I I_u + W_R R_u$, and calculating the scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W}.
\end{equation}
Thus, \emph{per capita} testing rate for compartment $Z$ is $F_Z=\sigma W_Z$, where $Z \in \{S,I,R\}$.
For a high-sensitivity test, infected people typically flow through to the ``confirmed positive'' ($I_c$/$R_c$) compartments and are thus unavailable for further testing.
Over the course of the epidemic, a fixed testing rate as specified in \eqref{sigma} can
(if large enough) exhaust the pool of people available for testing,
leading to a singularity when no one is left untested.
Although this phenomenon does not affect our analysis of $\Rnum$, it does sometimes occur in
numerical solutions of the time dynamics: we discuss an adjustment to the model that solves
this problem in the Appendix.

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population, exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \crefrange{eq1}{eq12}, assumes: (i) there is a single force of infection, $\Lambda$, across all susceptible pools \jd{This is de-clarifying here; let's drop it: that is, where a susceptible individual moves to does not depend on who infected it}, (ii) $\eta_c \leq \eta_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. For this analysis, we also assume a perfectly specific test,($p_S=0$). This last assumption 
%% The latter assumption, combined with the assumption that the total population was initially distributed between untested and awaiting for negative test, that is $S_u(0)+S_n(0)=N(0)$ \bmb{I don't think this is what we need: I think we need $S_p(0)=S_c(0)=0$? The previous assumption is awkward because we must have some infected individuals at time 0, and is more than we need (we could start with some $R$ without loss of generality)}, 
reduces the model to 10 equations (equations \ref{eq3} and \ref{eq4} are eliminated).
%% Specifically, when the population is at steady state and the number of infected people is negligible \bmb{do we really need anything other than the previously stated conditions?}, the assumption of perfect specificity results in $S_p=0$ in \cref{eq3} and $d S_c/dt=0$ from \cref{eq4}. Also, since we assumed $S_u+S_n=0$, $S_c=0$ and stays in this state.

% \newgeometry{margin=1cm} % modify this if you need even more space
% \begin{landscape}
\begin{table}[htp]
\centering
{%\tiny %
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & Per capita testing intensity & 1/day & 0.01 \\ \hline 
  $\eta_w$  & Relative probability of transmission for ``waiting'' individuals & - & - \\ \hline
  $\eta_c$  & Relative probability of transmission for ``confirmed positive'' individuals & - & -  \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $W_S, W_I, W_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline

  \end{tabular}
  }%
\caption{\label{tab:params} Model parameters for \crefrange{eq1}{eq12}.}
\end{table}
%\end{landscape}
%\restoregeometry

The Disease-Free Equilibrium (DFE) for the SIR model, \crefrange{eq1}{eq12}, is given by solving the coupled system including $S_u+S_n=N_0$ and $F_S S_u-\omega S_n=0$. The DFE is

@@

Further details are provided in the Appendix.

\todo{Something needs to be said here about simulation part.} The analytical calculation of the next generation matrix, $G$, and the derivation of $\Rnum$ was carried out in Maple (ref) by using simple linear Algebra package. 
  We used \Rlogo for the simulation part which included using the explicit expression of $\Rnum$ \ref{R0} as a function of the underlying parameters, specifying the parameters a realistic range, calculating the corresponding $\Rnum$ and plotting the contours of $\Rnum$ by using ggplot package in \Rlogo. The related result is presented in Figure \ref{pan}, where  panel \eqref{p.a} represents the random testing, and panel \eqref{p.b} is representing non-random testing. The parameter values or ranges are as follows (I like using a table here, with the param, their units and values). It is notable that in the simulation part, the explicit expression of $\Rnum$ was used and not the approximation \eqref{eq:R0appr}. This simulation reflects the behavior of $\Rnum$ with respect to the selected parameters for two different testing strategies: (i) random testing, represented by all testing weights to be the same, $W_S=W_I=W_R$, and (ii) non-random testing, when testing weight are not equal. For simulation purposes we chose $W_S=W_I=W_R=1$ for random testing, and $W_S=0.3$ and $W_I=W_R=1$ for non-random testing strategy. Note that the critical contour of $\Rnum=1$ is plotted in solid line in Figure \ref{pan}.

@@

% %%%%%%%
\section{Supplementary Material}

\subsection{Model and Calculation of $\Rnum$}

The model is 
\begin{align}
% \label{model}
 d S_u/dt &= -\Lambda S_u - F_S S_u + \omega S_n, \label{eq1}\\
 d S_n/dt &= -\Lambda S_n + (1-p_S) F_S S_u - \omega S_n, \label{eq2}\\
 d S_p/dt &= -\Lambda S_p + p_S F_S S_u - \omega S_p, \label{eq3}\\
 d S_c/dt &= -\Lambda S_c + \omega S_p, \label{eq4}\\
 d I_u/dt &= \Lambda S_u - F_I I_u + \omega I_n  - \gamma I_u,  \label{eq5}\\
 d I_n/dt &= \Lambda S_n + (1-p_I) F_I I_u - \omega I_n -\gamma I_n, \label{eq6}\\
 d I_p/dt &= \Lambda S_p + p_I F_I I_u - \omega I_p -\gamma I_p, \label{eq7}\\
 d I_c/dt &= \Lambda S_c + \omega I_p - \gamma I_c,  \label{eq8}\\
 d R_u/dt &= \gamma I_u - F_R R_u + \omega R_n, \label{eq9}\\
 d R_n/dt &= \gamma I_n + (1-p_R) F_R R_u - \omega R_n,  \label{eq10}\\
 d R_p/dt &= \gamma I_p + p_R F_R R_u  - \omega R_p,  \label{eq11}\\
 d R_c/dt&= \gamma I_c + \omega R_p,  \label{eq12}\\
 dN/dt &= \omega (S_n + I_n + R_n),   \label{eq13}\\
 dP/dt &= \omega(I_p + R_p) \label{eq14},
\end{align}

where $\beta$, (in units of 1/day), is the disease transmission rate, $\eta_w$ and $\eta_c$ are the isolation parameters for awaiting and reported individuals, respectively. $\Lambda$ is the force of infection and is defined as

\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I_u+\eta_w I_n+\eta_w I_p+ \eta_c I_c)}{N_0},
\end{equation}

% %

>>>>

\usepackage{xspace}

%\newcommand{\Rlogo}{\protect\includegraphics[height=2ex,keepaspectratio]{pix/Rlogo.pdf}\xspace}
% BMB: we don't need the weird graphical logo ...
\newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}

@@

We developed a deterministic model, Eqs. \eqref{model}, which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered (thus this is an SIR-type model), and testing status categorizes people as \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (Figure~\ref{fig:flowchart}). Two `accumulator' compartments, $N$ and $P$, were also incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of $\Rnum$ are presented in the Supplementary Materials.

\begin{figure}[!h] 
\begin{center} 
\includegraphics[scale=0.3]{./pix/sir_comp.png}
\caption{\small Flowchart of the SIR (Susceptible-Infectious-Recovered) model, \ref{model}. For the description of the parameters see \cref{tab:params}.
\label{fig:flowchart}}
\end{center} 
\end{figure}

Table~\ref{tab:params} defines the model parameters, which are generally straightforward
\percap flows between compartments, or modifiers to these flow rates.
The novel component of the model comes in through the compartment-specific relative testing weights $W_S$, $W_I$ and $W_R$; these give the relative rates at which people in the $S$, $I$, and $R$ compartments are tested, respectively. For example, $W_I/W_S=2$ means that infected individuals are tested at twice the \percap rate of susceptible individuals.

In order to link to more applied models, we constructed this model so we could specify the total \percap testing rate. We do this by defining the weighted size of the testing pool $W = W_S S_u + W_I I_u + W_R R_u$, and calculating the scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W}.
\end{equation}
Thus, \percap testing rate for compartment $Z$ is $F_Z=\sigma W_Z$, where $Z \in \{S,I,R\}$.
For a high-sensitivity test, infected people typically flow through to the ``confirmed positive'' ($I_c$/$R_c$) compartments and are thus unavailable for further testing.
Over the course of the epidemic, a fixed testing rate as specified in \eqref{sigma} can
(if large enough) exhaust the pool of people available for testing,
leading to a singularity when no one is left untested.
Although this phenomenon does not affect our analysis of $\Rnum$, it does sometimes occur in
numerical solutions of the time dynamics: we discuss an adjustment to the model that solves
this problem in the Supplementary material.

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population, exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection, $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I_u+\eta_w I_n+\eta_w I_p+ \eta_c I_c)}{N_0},
\end{equation}
across all susceptible pools (see Table~\ref{tab:params} for detail), (ii) $\eta_c \leq \eta_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. For this analysis, we also assume a perfectly specific test,($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S_p(0)=S_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & Per capita testing intensity & 1/day & 0.01 \\ \hline 
  $\eta_w$  & Relative probability of transmission for ``waiting'' individuals & - & - \\ \hline
  $\eta_c$  & Relative probability of transmission for ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.39 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $W_S, W_I, W_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.}
\end{table}

The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by solving the coupled system including $S_u+S_n=N_0$ and $F_S S_u-\omega S_n=0$. The DFE is

@@

Further details are provided in the Supplementary material section.

The analytical calculation of the next generation matrix, $G$, and the derivation of $\Rnum$ was carried out in Maple (ref?) by using simple linear Algebra package. 
  We used R for the simulation part which included using the explicit expression of $\Rnum$ \ref{R0} as a function of the underlying parameters, specifying the parameters a realistic range, calculating the corresponding $\Rnum$ and plotting the contours of $\Rnum$ by using ggplot package in R. The related result is presented in Figure \ref{pan}, where  panel \eqref{p.a} represents the random testing, and panel \eqref{p.b} is representing non-random testing. The parameter values or ranges are as follows (I like using a table here, with the param, their units and values). It is notable that in the simulation part, the explicit expression of $\Rnum$ was used and not the approximation \eqref{eq:R0appr}. This simulation reflects the behavior of $\Rnum$ with respect to the selected parameters for two different testing strategies: (i) random testing, represented by all testing weights to be the same, $W_S=W_I=W_R$, and (ii) non-random testing, when testing weight are not equal. For simulation purposes we chose $W_S=W_I=W_R=1$ for random testing, and $W_S=0.3$ and $W_I=W_R=1$ for non-random testing strategy. Note that the critical contour of $\Rnum=1$ is plotted in solid line in Figure \ref{pan}.

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
% \widetext
\begin{center}
\textbf{\large Supplemental Materials: Title for main text}
\end{center}
%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{page}{1}
\makeatletter
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\bibnumfmt}[1]{[S#1]}
\renewcommand{\citenumfont}[1]{S#1}
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%

\subsection{Model and calculation of $\Rnum$}

The model is 
\begin{subequations}\label{model}
\begin{align}
 d S_u/dt &= -\Lambda S_u - F_S S_u + \omega S_n, \\
 d S_n/dt &= -\Lambda S_n + (1-p_S) F_S S_u - \omega S_n, \\
 d S_p/dt &= -\Lambda S_p + p_S F_S S_u - \omega S_p, \\
 d S_c/dt &= -\Lambda S_c + \omega S_p, \\
 d I_u/dt &= \Lambda S_u - F_I I_u + \omega I_n  - \gamma I_u, \\
 d I_n/dt &= \Lambda S_n + (1-p_I) F_I I_u - \omega I_n -\gamma I_n, \\
 d I_p/dt &= \Lambda S_p + p_I F_I I_u - \omega I_p -\gamma I_p, \\
 d I_c/dt &= \Lambda S_c + \omega I_p - \gamma I_c, \\
 d R_u/dt &= \gamma I_u - F_R R_u + \omega R_n, \\
 d R_n/dt &= \gamma I_n + (1-p_R) F_R R_u - \omega R_n, \\
 d R_p/dt &= \gamma I_p + p_R F_R R_u  - \omega R_p, \\
 d R_c/dt&= \gamma I_c + \omega R_p, \\
 dN/dt &= \omega (S_n + I_n + R_n),  \\
 dP/dt &= \omega(I_p + R_p) ,
\end{align}
\end{subequations}

where $\beta$, (in units of 1/day), is the disease transmission rate, $\eta_w$ and $\eta_c$ are the isolation parameters for awaiting and reported individuals, respectively. $\Lambda$ is the force of infection defined in Eq. \eqref{Lambda}.
